---
title: "dplyr"
author: "Mohammad Wasiq"
date: "10/12/2021"
output: html_document
---

# The Tidyverse
The *tidyverse* is a collection of R Packages specifically designed for Data Science. The following tutorials explain how to use various functions in these packages.

## dplyr 
A package designed for **manipulating data.**

### Arrange Rows in R
<br>**arrange()**
<br>1. **Arrange by One Column**
<br>2. **Arrange by Multiple Columns**
<br>3. **Arrange Rows in a Custom order**

```{r a1}
library(dplyr)
#create data frame
df <- data.frame(player = c('A', 'B', 'C', 'D', 'E', 'F', 'G'),
                 points = c(12, 14, 14, 15, 20, 18, 29),
                 assists = c(3, 5, 7, 8, 14, NA, 9))

#view data frame 
df

# 1: Arrange by One Column
df %>% arrange(points)

# To sort in descending order, we can use the desc() function:
df %>% arrange(desc(points))

df %>% arrange(assists)

# 2: Arrange by Multiple Columns
# sort by points, then assists 
df %>% arrange(points , assists)


# 3: Arrange Rows in a Custom order
#sort by player with custom order
df %>% arrange(factor(player, levels = c('D', 'C', 'A', 'B', 'E', 'F', 'G')))
```

### Remove Rows
1. Remove any row with NA’s
<br>**df %>%
<br>  na.omit()**
2. Remove any row with NA’s in specific column
<br>**df %>%
<br>  filter(!is.na(column_name))**
3. Remove duplicates
<br>**df %>%
<br>  distinct()**
4. Remove rows by index position
<br>**df %>%
<br>  filter(!row_number() %in% c(1, 2, 4))**
5. Remove rows based on condition
<br>**df %>%
<br>  filter(column1=='A' | column2 > 8)**

```{r remove1}
#create data frame
df <- data.frame(team=c('A', 'A', 'B', 'B', 'C', 'C'),
                 points=c(4, NA, 7, 5, 9, 9),
                 assists=c(1, 3, 5, NA, 2, 2))

#view data frame
df

#1: Remove Any Row with NA’s
df %>%
na.omit()

# 2: Remove Any Row with NA’s in Specific Columns
# remove any row with NA in 'points' column:
df %>%
filter(!is.na(points))

# 3: Remove Duplicate Rows
df %>%
distinct()

# 4: Remove Rows by Index Position
# Remove rows 1,2 and 4
df %>%
filter(!row_number() %in% c(1,2,4))

# 5: Remove Rows Based on Condition
#only keep rows where team is equal to 'A' or points is greater than 8
df %>%
filter(team=='A' | points > 8)
```

### Remove Columns
<br>**select()**
<br>1. **Remove Columns by Name**
<br>2. **Remove Columns in List**
<br>3. **Remove Columns in Range**
<br>4. **Remove Columns that Contain a Phrase**
<br>5. **Remove Columns that Start with Certain Letters**
<br>6. **Remove Columns that End with Certain Letters**
<br>7. **Remove Columns that End with Certain Letters**

```{r select1}
#create data frame
df <- data.frame(player = c('a', 'b', 'c', 'd', 'e'),
                 position = c('G', 'F', 'F', 'G', 'G'),
                 points = c(12, 15, 19, 22, 32),
                 rebounds = c(5, 7, 7, 12, 11))

#view data frame
df

# 1: Remove Columns by Name
#remove column named 'points'
df %>% select(-points)

# 2: Remove Columns in List
#remove columns named 'points' or 'rebounds'
df %>% select(-one_of('points' , 'rebounds'))

# 3: Remove Columns in Range
#remove columns in range from 'position' to 'rebounds'
df %>% select(-(position : rebounds))

# 4: Remove Columns that Contain a Phrase
#remove columns that contain the word 'points'
df %>% select(-contains('points'))

#  5: Remove Columns that Start with Certain Letters
#remove columns that start with 'po'
df %>% select(-starts_with('po'))

# 6: Remove Columns that End with Certain Letters
#remove columns that end with 's'
df %>% select(-ends_with('s'))

# 7: Remove Columns by Position
#remove columns in position 1 and 4
df %>% select(-1 , -4)
```

### Add Columns to Data Frame in R 
1. Add Column at End of Data Frame
<br>**df %>%
<br>  mutate(new_col=c(1, 3, 3, 5, 4))**
2. Add Column Before Specific Column
<br>**df %>%
<br>mutate(new_col=c(1, 3, 3, 5, 4),
<br>        .before=col_name)**
3. Add Column After Specific Column
<br>**df %>%
<br>mutate(new_col=c(1, 3, 3, 5, 4),
<br>        .after=col_name)**
4. Add Column Based on Other Columns
<br>df %>%
<br>  mutate(new_col= if_else(.$col_name > 10, 'A', 'B'))

```{r add1}
#create data frame
df <- data.frame(team=c('A', 'A', 'A', 'A', 'B', 'B', 'B', 'B'),
                 points=c(12, 14, 19, 24, 24, 22, 30, 9),
                 assists=c(4, 6, 6, 8, 3, 7, 8, 11))

#view data frame
df

# 1: Add Column at End of Data Frame
#add 'blocks' column at end of data frame
df <- df %>%
        mutate(blocks=c(1, 3, 3, 2, 4, 3, 6, 2))

#view data frame
df

# add an empty column by simply assigning NA to every value in the new column
df = df %>% 
mutate(blocks=NA)

# View Data Frame
df

# 2: Add Column Before Specific Column
#add 'blocks' column before 'points' column
df <- df %>%
        mutate(blocks=c(1, 3, 3, 2, 4, 3, 6, 2),
               .before=points)
#view data frame
df

# 3. Add Column After Specific Column
#add 'blocks' column after 'points' column
df <- df %>%
        mutate(blocks=c(1, 3, 3, 2, 4, 3, 6, 2),
              .after=points)

#view data frame
df

# 4: Add Column Based on Other Columns
#add 'status' column whose values depend on value in 'points' column
df <- df %>%
        mutate(status= if_else(.$points > 20, 'Good', 'Bad'))

#view data frame
df
```

### Count Observations by Group 
<br>**count()**
<br>1. **Count by One Variable**
<br>2. **Count by Multiple Variables**
<br>3. **Weighted Count**

```{r count1}
#create data frame
df <- data.frame(team = c('A', 'A', 'A', 'B', 'B', 'B', 'B', 'B', 'C', 'C', 'C', 'C'),
                 position = c('G', 'G', 'F', 'G', 'F', 'F', 'F', 'G', 'G', 'F', 'F', 'F'),
                 points = c(4, 13, 7, 8, 15, 15, 17, 9, 21, 22, 25, 31))

#view data frame
df

# 1 Count by One Variable
#count total observations by variable 'team'
df %>% count(team)

# we can also sort the counts if we’d like:
#count total observations by variable 'team'
df %>% count(team, sort=TRUE)

# 2. Count by Multiple Variables
#count total observations by 'team' and 'position'
df %>% count(team, position)

#  3: Weighted Count
df %>% count(team, wt=points)
```


### Count Distinct Values
<br>**n_distinct()**
<br> 1. **n_distinct(df$column_name)** : Count Distinct Values in One Column
<br> 2. **sapply(df, function(x) n_distinct(x))** : Count Distinct Values in All Columns
<br> 3. **df %>%**
<br>  **group_by(grouping_column) %>%**
<br>  **summarize(count_distinct = n_distinct(values_column))** : Count Distinct Values by Group

```{r ndistinct1}
#create data frame
df <- data.frame(team=c('A', 'A', 'A', 'A', 'B', 'B', 'B', 'B'),
                 points=c(6, 6, 8, 10, 9, 9, 12, 12),
                 assists=c(3, 6, 4, 2, 4, 5, 5, 9))

#view data frame
df

# 1: Count Distinct Values in One Column

#count distinct values in 'team' column
n_distinct(df$team)

# 2: Count Distinct Values in All Columns

#count distinct values in every column
sapply(df, function(x) n_distinct(x))

# 3 Count Distinct Values by Group

# count distinct 'points' values by 'team'
df %>%
  group_by(team) %>%
  summarise(distinct_points = n_distinct(points))
```


### Create a Crosstab
<br>*library(dplyr)*
<br>*library(tidyr)*
<br>**df %>%**
<br>  **group_by(var1, var2) %>%**
<br>  **tally() %>%**
<br>  **spread(var1, n)**

```{r ct1}
#create data frame
df <- data.frame(team=c('A', 'A', 'A', 'A', 'B', 'B', 'B', 'B'),
                 position=c('G', 'G', 'F', 'C', 'G', 'F', 'F', 'C'),
                 points=c(7, 7, 8, 11, 13, 15, 19, 13))

#view data frame
df

# create a crosstab for the ‘team’ and ‘position’ variables:

library(dplyr)
library(tidyr)

#produce crosstab 
df %>%
  group_by(team, position) %>%
  tally() %>%
  spread(team, n)

# switch the rows and columns of the crosstab by switching the variable used in the spread() function:

#produce crosstab with 'position' along columns
df%>%
  group_by(team, position) %>%
  tally() %>%
  spread(position, n)
```


### Filter Rows
<br>**filter()**

<br>1. **Filter Rows Equal to Some Value**
<br>2. **Filter Rows Using ‘And’ (&)**
<br>3. **Filter Rows Using ‘Or’ (|)**
<br>4. **Filter Rows with Values in a List**
<br>5. **Filter Rows Using Less Than or Greater Than**

```{r filter1}
#view first six rows of starwars dataset
head(starwars)
dim(starwars)

# 1. Filter Rows Equal to Some Value
# filter the row eqaul to Droid
starwars %>% filter(species == 'Droid')

# 2. Filter Rows Using ‘And’ (&)
#  filter for rows where the species is Droid and the eye color is red:
starwars %>% filter(species == 'Droid' & eye_color == 'red')

# 3. Filter Rows Using ‘Or’
# filter for rows where the species is Droid or the eye color is red:
starwars %>% filter(species == 'Droid' | eye_color == 'red')

# 4. Filter Rows with Values in a List
# filter for rows where the eye color is in a list of colors:
# colors are blue, yellow, red
starwars %>% filter(eye_color %in% c('blue', 'yellow', 'red'))

# 5. Filter Rows Using Less Than or Greater Than
# filter rows using less than or greater than operations on numeric variables
#find rows where height is greater than 250
starwars %>% filter(height > 250)

#find rows where height is between 200 and 220
starwars %>% filter(height > 200 & height < 220)

#find rows where height is above the average height
starwars %>% filter(height > mean(height, na.rm = TRUE))
```


### Filter by Multiple Conditions
<br>*1. Filter by Multiple Conditions Using OR (|)*
<br>**library(dplyr)**
<br>**df %>%**
<br>**filter(col1 == 'A' | col2 > 90)**
<br>*2. Filter by Multiple Conditions Using AND (&)*
<br>**library(dplyr)**
<br>**df %>%**
<br>**filter(col1 == 'A' & col2 > 90)**

```{r mc1}
#create data frame
df <- data.frame(team=c('A', 'A', 'B', 'B', 'C'),
                 points=c(99, 90, 86, 88, 95),
                 assists=c(33, 28, 31, 39, 34),
                 rebounds=c(30, 28, 24, 24, 28))

#view data frame
df

# 1. Filter by Multiple Conditions Using OR
# filter for rows where team is equal to 'A' or points is greater than 90
df %>% 
filter(team == 'A' | points> 90)

#filter for rows where team is equal to 'A' or 'C' or points is less than 89
df %>%
  filter(team == 'A' | team == 'C' | points > 90)

# 2. Filter by Multiple Conditions Using AND
#filter for rows where team is equal to 'A' or 'C' or points is less than 89
df %>% filter(team == 'A' & points > 90)

#filter where team is equal to 'A' and points > 89 and assists < 30
df %>% filter(team == 'A' & points>89 & assists<30)
```


### Use a “not in” Filter
<br>**df %>%**
<br>  **filter(!col_name %in% c('value1', 'value2', 'value3', ...))**
  
<br>1. **Filter for Rows that Do Not Contain Value in One Column**
<br>2. **Filter for Rows that Do Not Contain Value in Multiple Columns**

```{r not1}
#create data frame
df <- data.frame(team=c('A', 'A', 'B', 'B', 'C', 'C', 'D', 'D'),
                 position=c('G', 'G', 'F', 'G', 'F', 'C', 'C', 'C'),
                 points=c(12, 14, 19, 24, 36, 41, 18, 29))

#view data frame
df

#filter for rows where team name is not 'A' or 'B'
df %>%
  filter(!team %in% c('A', 'B'))

# 2: Filter for Rows that Do Not Contain Value in Multiple Columns
# #filter for rows where team name is not 'A' and position is not 'C'
df %>%
  filter(!team %in% c('A') & !position %in% c('C'))
```


### Remove Rows with NA Values
<br>1. *Remove rows with NA value in any column*
<br>**df %>%**
<br>  **na.omit()**
<br>2. *Remove Rows with NA Values in Certain Columns*
<br>**df %>%**
<br>  **filter_at(vars(col1, col2), all_vars(!is.na(.)))**
<br>3. *Remove Rows with NA Values in One Specific Column*
<br>**df %>%**
<br>  **filter(!is.na(col1))**

```{r rna1}
#create data frame with some missing values
df <- data.frame(team=c('A', 'A', 'B', 'B', 'C'),
                 points=c(99, 90, 86, 88, NA),
                 assists=c(33, NA, 31, 39, 34),
                 rebounds=c(NA, 28, 24, 24, 28))

#view data frame
df

# 1. Remove rows with NA value in any column
df %>%
  na.omit()

# #remove rows with NA value in any column
df %>%
  filter_at(vars(points, assists), all_vars(!is.na(.)))

# 3: Remove Rows with NA Values in One Specific Column
#remove rows with NA value in 'points' column
df %>%
  filter(!is.na(points))
```

### Recode Values
<br>**recode()**
<br>1. **Recode a Single Column in a Dataframe**
<br>2.**Recode a Single Column in a Dataframe and Provide NA Values**
<br>3. **Recode Multiple Columns in a Dataframe**

```{r rval1}
#create dataframe 
df <- data.frame(player = c('A', 'B', 'C', 'D'),
                 points = c(24, 29, 13, 15),
                 result = c('Win', 'Loss', 'Win', 'Loss'))

#view dataframe 
df

#1. change 'Win' and 'Loss' to '1' and '0'
df %>%
  mutate(result=recode(result,'Win'='1', 'Loss'='0'))

# 2.Recode a Single Column in a Dataframe and Provide NA Values
#change 'Win' to '1' and give all other values a value of NA
df %>% 
  mutate(result=recode(result, 'Win'='1', .default=NA_character_))

# 3: Recode Multiple Columns in a Dataframe
#recode 'player' and 'result' columns
df %>%
  mutate(player = recode(player, 'A' = 'Z'),
         result = recode(result, 'Win'='1' , 'Loss' = '0'))
```


### Reorder Columns
<br>**select()**
<br>1. **Move a Column to the First Position**
<br>2. **Move a Column to the Last Position**
<br>3. **Reorder Multiple Columns**
<br>4. **Reorder Columns Alphabetically**
<br>5. **Reverse Column Order**

```{r reorder1}
#create data frame
df <- data.frame(player = c('a', 'b', 'c', 'd', 'e'),
                 position = c('G', 'F', 'F', 'G', 'G'),
                 points = c(12, 15, 19, 22, 32),
                 rebounds = c(5, 7, 7, 12, 11))

#view data frame
df

# 1: Move a Column to the First Position
#move column 'points' to first position
df %>% select(points, everything())

# 2: Move a Column to the Last Position
#move column 'points' to last position
df %>% select(-points, points)

# 3. Reorder Multiple Columns
#change all column names to uppercase
df %>% select(rebounds, position, points, player)

# 4. Reorder Columns Alphabetically
#order columns alphabetically
df %>% select(order(colnames(.)))

# 5. Reverse Column Order
#reverse column order
df %>% select(rebounds : player , everything())
```


### Rename Column by Index Position 
<br>1. *Rename One Column by Index*
<br>**df %>%**
<br>   **rename(new_name1 = 1)**
<br>2. *Rename Multiple Columns by Index*
<br>**df %>%**
<br>  **rename(new_name1 = 1,**
<br>        **new_name2 = 2,**
<br>        **new_name3 = 3)**


```{r renmae1}
#create data frame
df <- data.frame(team=c('A', 'A', 'A', 'A', 'B', 'B', 'B', 'B'),
                 points=c(12, 14, 19, 24, 24, 22, 30, 9),
                 assists=c(4, 6, 6, 8, 3, 7, 8, 11))

df
#1. rename column in index position 1
df <- df %>%
        rename(team_new = 1)

#view updated data frame
df

# 2. Rename Multiple Columns by Index
#rename column in index position 1
df<- df %>%
       rename(points_new = 2,
              assists_new = 3)

#view updated data frame
df
```


### Group & Summarize Data
<br>**data %>%**
<br>  **group_by(col_name) %>%**
<br>  **summarize(summary_name = summary_function)**
<br>1. *Find Mean & Median by Group*
<br>2. *Find Measures of Spread by Group*
<br>3. *Find Count by Group*
<br>4. *Find Percentile by Group*

```{r grp1}
#obtain rows and columns of mtcars
dim(mtcars)
#view first six rows of mtcars
head(mtcars)

# 1: Find Mean & Median by Group
#find mean mpg by cylinder
mtcars %>%
  group_by(cyl) %>%
  summarise(mean_mpg = mean(mpg, na.rm = TRUE))

# find the median mpg by cyl
mtcars %>%
  group_by(cyl) %>%
  summarise(median_mpg = median(mpg, na.rm = TRUE))

# 2. Find Measures of Spread by Group
#find sd, IQR, and mad by cylinder
mtcars %>%
  group_by(cyl) %>%
  summarize(sd_mpg = sd(mpg, na.rm = TRUE),
            iqr_mpg = IQR(mpg, na.rm = TRUE),
            mad_mpg = mad(mpg, na.rm = TRUE))

# 3: Find Count by Group
#find row count and unique row count by cylinder
mtcars %>% 
  group_by(cyl) %>%
  summarise(count_mpg = n(),
            u_count_mpg = n_distinct(mpg))

# 4: Find Percentile by Group
#find 90th percentile of mpg for each cylinder group
mtcars %>%
  group_by(cyl) %>%
  summarize(quant90 = quantile(mpg, probs = 0.9))
```


### Maximum Value by Group
<br>1. **Maximum Value by Group**
<br>2. **Return Rows that Contains Max Value by Group**
<br>3. **Return a Single Row that Contains Max Value by Group**

```{r max1}
#create data frame
df <- data.frame(team = c('A', 'A', 'A', 'B', 'B', 'B', 'B'),
                 position = c('G', 'F', 'F', 'G', 'G', 'G', 'F'),
                 points = c(12, 15, 19, 22, 34, 34, 39))

#view data frame
df

# 1. Find Max Value by Group
#find max value by team and position
df %>%
  group_by(team, position) %>%
  summarise(max = max(points, na.rm=TRUE))

# 2: Return Rows that Contains Max Value by Group
#find rows that contain max points by team and position
df %>%
  group_by(team, position) %>%
  filter(points == max(points, na.rm=TRUE))

# 3: Return a Single Row that Contains Max Value by Group
#find rows that contain max points by team and position
df %>%
  group_by(team, position) %>%
  slice(which.max(points))
```


### Replace NA with Zero
<br>1.**df <- df %>% replace(is.na(.), 0)** : Replace all NA values with zero
<br>2. **df %>% mutate(col1 = ifelse(is.na(col1), 0, col1))** : Replace NA values with zero in column named col1
<br>3. *Replace NA values with zero in columns col1 and col2*
<br>**df %>% mutate(col1 = ifelse(is.na(col1), 0, col1),**
<br>              **col2 = ifelse(is.na(col2), 0, col2))**


```{r rwz1}
library(dplyr)
#create data frame
df <- data.frame(player=c('A', 'B', 'C', 'D', 'E'),
                 pts=c(17, 12, NA, 9, 25),
                 rebs=c(3, 3, NA, NA, 8),
                 blocks=c(1, 1, 2, 4, NA))

#view data frame
df

# 1: Replace All NA Values in All Columns
df <- df %>% replace(is.na(.), 0)
df

# 2: Replace NA Values in a Specific Column
#replace NA values with zero in rebs column only
df<- df %>% mutate(rebs= ifelse(is.na(rebs), 0, rebs))
df

# 3: Replace NA Values in One of Several Columns
#replace NA values with zero in rebs and pts columns
df <- df %>% mutate(rebs = ifelse(is.na(rebs), 0, rebs),
                    pts = ifelse(is.na(pts), 0, pts))

df
```


### Join Multiple Data Frames
<br>**left_join()**

```{r join1}
#create data frame
df1 <- data.frame(a = c('a', 'b', 'c', 'd', 'e', 'f'),
                  b = c(12, 14, 14, 18, 22, 23))

df2 <- data.frame(a = c('a', 'a', 'a', 'b', 'b', 'b'),
                  c = c(23, 24, 33, 34, 37, 41))

df3 <- data.frame(a = c('d', 'e', 'f', 'g', 'h', 'i'),
                  d = c(23, 24, 33, 34, 37, 41))

# To join all three data frames together
df1 %>%
  left_join(df2 , by= 'a') %>%
  left_join(df3 , by= 'a')

#join the three data frames and save result as new data frame named all_data

all_data <- df1 %>%
  left_join(df2, by= 'a') %>%
  left_join(df3, by= 'a')
#view summary of resulting data frame 
glimpse(all_data)  
```


### Anti_join to Find Unmatched Records
<br>Return all rows in one data frame that do not have matching values in another data frame.
<br>**anti_join()**
<br>**anti_join(df1, df2, by='col_name')**
<br>1. *Use anti_join() with One Column*
<br>2. *Use anti_join() with Multiple Columns*

```{r join2}
#create data frames
# 1. Use anti_join() with One Column
df1 <- data.frame(team=c('A', 'B', 'C', 'D', 'E'),
                  points=c(12, 14, 19, 24, 36))

df2 <- data.frame(team=c('A', 'B', 'C', 'F', 'G'),
                  points=c(12, 14, 19, 33, 17))
#perform anti join using 'team' column
anti_join(df1, df2, by= 'team')

# 2. Use anti_join() with Multiple Columns

#create data frames
df1 <- data.frame(team=c('A', 'A', 'A', 'B', 'B', 'B'),
                  position=c('G', 'G', 'F', 'G', 'F', 'C'),
                  points=c(12, 14, 19, 24, 36, 41))

df2 <- data.frame(team=c('A', 'A', 'A', 'B', 'B', 'B'),
                  position=c('G', 'G', 'C', 'G', 'F', 'F'),
                  points=c(12, 14, 19, 33, 17, 22))

#perform anti join using 'team' and 'position' columns
anti_join(df1, df2, by = c('team', 'position'))
```


### case_when()
<br>Used to create new variables from existing variables.
<br>**df %>%**
<br>  **mutate(new_var = case_when(var1 < 15 ~ 'low',**
<br>                             **var2 < 25 ~ 'med',**
<br>                             **TRUE ~ 'high'))**

<br>1. *Create New Variable from One Existing Variable*
<br>2. *Create New Variable from Multiple Variables*

```{r cw1}
#create data frame
df <- data.frame(player = c('AJ', 'Bob', 'Chad', 'Dan', 'Eric', 'Frank'),
                 position = c('G', 'F', 'F', 'G', 'C', NA),
                 points = c(12, 15, 19, 22, 32, NA),
                 assists = c(5, 7, 7, 12, 11, NA))

#view data frame
df

# 1: Create New Variable from One Existing Variable
# create a new variable called quality whose values are derived from the points column
# Here is exactly how the case_when() function created the values for the new column:
# i. If the value in the points column is greater than 20, then the value in the quality column is “high”
# ii. Else, if the value in the points column is greater than 15, then the value in the quality column is “med”
# iii. Else, if the value in the points column is less than or equal to 15 (or a missing value like NA), then the value in the quality column is “low”
# iv. Else, if the value in the points column is greater than 15, then the value in the quality column is “med”
df %>%
  mutate(quality= case_when(points > 20 ~ 'high',
                            points > 15~ 'med',
                            TRUE~ 'low'))

# 2: Create New Variable from Multiple Variables
# create a new variable called quality whose values are derived from both the points and assists column
df %>%
  mutate(quality = case_when(is.na(points) ~ 'missing',
                             points > 15 & assists > 10 ~ 'great',
                             points > 15 & assists > 5 ~ 'good',
                             TRUE ~ 'average' ))
```


### Use bind_rows and bind_cols
<br>To bind together two data frames by their rows:
<br>**bind_rows(df1, df2, df3, ...)**

To bind together two data frames by their columns:
<br>**bind_cols(df1, df2, df3, ...)**

```{r bind1}
#create data frames
df1 <- data.frame(team=c('A', 'A', 'B', 'B'),
                  points=c(12, 14, 19, 24))


df2 <- data.frame(team=c('A', 'B', 'C', 'C'),
                  points=c(8, 17, 22, 25))

df3 <- data.frame(team=c('A', 'B', 'C', 'C'),
                  assists=c(4, 9, 12, 6))

#row bind together data frames
bind_rows(df1, df2, df3)

# Column bind togrther data franes
bind_cols(df1, df2, df3)
```

### Select Columns by Index
<br>*Select columns in specific index positions*
<br>**df %>%**
<br>  **select(1, 4, 5)**
<br>*Exclude columns in specific index positions*
<br>**df %>%**
<br>  **select(-c(1,2))**

```{r scbi1}
#create data frame
df <- data.frame(team=c('A', 'B', 'C', 'D', 'E'),
                 points=c(99, 90, 86, 88, 95),
                 assists=c(33, 28, 31, 39, 34),
                 rebounds=c(30, 28, 24, 24, 28),
                 blocks=c(14, 19, 22, 18, 15))

#view data frame
df

# 1: Select Columns in Specific Index Positions
#select columns in position 1, 4, and 5
df %>% 
  select(1,4,5)

#2: Select Columns in Range
#select columns in position 2 through 4
df %>%
  select(2:4)

# 3: Exclude Specific Columns
#select all columns except those in position 1 and 2
df %>%
  select(-c(1, 2))
```


### Calculate Relative Frequencies
<br>1. **Relative Frequency of One Variable**
<br>2. **Relative Frequency of Multiple Variables**
<br>3. **Display Relative Frequencies as Percentages**

```{r rf1}
#create data frame
df <- data.frame(team = c('A', 'A', 'A', 'B', 'B', 'B', 'B'),
                 position = c('G', 'F', 'F', 'G', 'G', 'G', 'F'),
                 points = c(12, 15, 19, 22, 32, 34, 39))

#view data frame
df

# 1: Relative Frequency of One Variable
# To calculate the relative frequency of each team in the data frame
df %>% 
  group_by(team) %>% 
  summarize(n= n()) %>%
  mutate(rel_freq = n / sum(n))

#  2: Relative Frequency of Multiple Variables
# To calculate the relative frequency of positions by team:
df %>% 
  group_by(team, position) %>%
  summarise(n= n()) %>%
  mutate(freq= n / sum(n))

# 3: Display Relative Frequencies as Percentages
# To calculate the relative frequency of positions by team and how to display these relative frequencies as percentages
df %>%
  group_by(team, position) %>%
  summarise(n= n()) %>%
  mutate(freq = paste0(round(100 * n/sum(n), 0), '%'))
```

### Frequency Table by Group
<br>**df %>%**
<br>  **group_by(var1, var2) %>%**
<br>  **summarize(Freq=n())**

```{r ftbg1}
#create data frame
df <- data.frame(team=c('A', 'A', 'A', 'A', 'B', 'B', 'B', 'B'),
                 position=c('G', 'G', 'G', 'F', 'G', 'F', 'F', 'C'))

#view data frame
df

#  create a frequency table that shows the frequency of each position, grouped by team.
df %>% 
  group_by(team, position) %>%
  summarise(Freq = n())
```


### Select the First Row by Group
<br>**df %>%**
<br>  **group_by(group_var) %>%**
<br>  **arrange(values_var) %>%**
<br>  **filter(row_number()==1)**

```{r frg1}
library(dplyr)
#create dataset
df <- data.frame(team=c('A', 'A', 'A', 'B', 'B', 'B', 'C', 'C', 'C', 'C'),
                 points=c(4, 9, 7, 7, 6, 13, 8, 8, 4, 17))

#view dataset
df

#  select the first row by group 
df %>% 
  group_by(team) %>%
  arrange(points) %>%
  filter(row_number()==1)

# By default, arrange() sorts the values in ascending order but we can easily sort the values in descending order instead
df %>% 
  group_by(team) %>%
  arrange(desc(points)) %>%
  filter(row_number()==1)

# select the 2nd row by group
df %>%
  group_by(team) %>%
  arrange(desc(points)) %>%
  filter(row_number()==2)

# select the last row by group:

df %>%
  group_by(team) %>%
  arrange(desc(points)) %>%
  filter(row_number()==n())
```


### Rank Variables by Group

<br>**df %>% arrange(group_var, numeric_var) %>%**
<br>    **group_by(group_var) %>%**
<br>    **mutate(rank = rank(numeric_var))**

1. **Rank in Ascending Order**
<br>2. **Rank in Descending Order**
<br>3. **How to Handle Ties in Ranking**


```{r rvbg1}
#create data frame
df <- data.frame(team = c('A', 'A', 'A', 'A', 'B', 'B', 'B', 'C', 'C', 'C'),
                 points = c(12, 28, 19, 22, 32, 45, 22, 28, 13, 19),
                 rebounds = c(5, 7, 7, 12, 11, 4, 10, 7, 8, 8))

#view data frame
df

#  1: Rank in Ascending Order
#  rank the points scored by players in ascending order, grouped by team
#rank points scored, grouped by team
df %>%
  arrange(team, points) %>%
  group_by(team) %>%
  mutate(rank = rank(points))

#  2: Rank in Descending Order
#rank points scored in reverse, grouped by team
df %>%
  arrange(team, points) %>%
  group_by(team) %>%
  mutate(rank = rank(-points))
```


### Mutate to Create New Variables
<br>1.**mutate()** – adds new variables to a data frame while preserving existing variables
<br>2. **transmute()** – adds new variables to a data frame and drops existing variables
<br>3. **mutate_all()** –  modifies all of the variables in a data frame at once
<br>4. **mutate_at()** –  modifies specific variables by name
<br>5. **mutate_if()** – modifies all variables that meet a certain condition


#### 1. mutate()
<br>The *mutate()* function adds new variables to a data frame while preserving any existing variables. The basic synax for mutate() is as follows:
<br>**data <- mutate(new_variable = existing_variable/3)**
<br>*data:* the new data frame to assign the new variables to
<br>*new_variable:* the name of the new variable
<br>*existing_variable:* the existing variable in the data frame that you wish to perform some operation on to create the new variable

```{r mut1}
#define data frame as the first six lines of the iris dataset
data <- head(iris)

#view data
data

#define new column root_sepal_width as the square root of the Sepal.Width variable
data %>%
  mutate(root_sepal_width = sqrt(Sepal.Width))

```

#### 2. transmute()
<br>The *transmute()* function adds new variables to a data frame and drops existing variables

```{r tm1}
#define data frame as the first six lines of the iris dataset
data <- head(iris)

#view data
data

#define two new variables and remove all existing variables
data %>%
  transmute(root_sepal_width = sqrt(Sepal.Width),
            root_petal_width = sqrt(Petal.Width))
```


#### 3. mutate_all()
<br>The *mutate_all()* function modifies all of the variables in a data frame at once, allowing you to perform a specific function on all of the variables by using the *funs()* function. 

```{r muteall1}
#define new data frame as the first six rows of iris without the Species variable
data2<- head(iris) %>%
  select(-Species)

data2

#divide all variables in the data frame by 10
data2 %>% mutate_all(funs(./10))

# Note that additional variables can be added to the data frame by specifying a new name to be appended to the old variable name:
data2 %>% mutate_all(funs(mod = ./10))
```

#### 4. mutate_at()
<br>The *mutate_at()* function modifies specific variables by name.

```{r muteat1}
# The following code illustrates how to divide two specific variables by 10 using mutate_at():
data2 %>%
  mutate_at(c('Sepal.Length', 'Sepal.Width'), funs(mod = ./10))
```

#### 5. mutate_if()
<br>The *mutate_if()* function modifies all variables that meet a certain condition. 

```{r muteif1}
# The following code illustrates how to use the mutate_if() function to convert any variables of type factor to type character:
#find variable type of each variable in a data frame
data <- head(iris)
sapply(data, class)

#convert any variable of type factor to type character
new_data <- data %>% 
  mutate_if(is.factor, as.character)
sapply(new_data, class)

# round any variables of type numeric to one decimal place:

#define data as first six rows of iris dataset
data <- head(iris)

#view data
data

#round any variables of type numeric to one decimal place
data %>% mutate_if(is.numeric, round, digits = 0)
```


### Create New Variables
<br>Create a new variable in a data frame using the *mutate()* and *case_when()*.

```{r cnw1}
#create data frame
df <- data.frame(player = c('a', 'b', 'c', 'd', 'e'),
                 position = c('G', 'F', 'F', 'G', 'G'),
                 points = c(12, 15, 19, 22, 32),
                 rebounds = c(5, 7, 7, 12, 11))

#view data frame
df

#  1: Create New Variable Based on One Existing Variable
#define new variable 'scorer' using mutate() and case_when()
df %>%
  mutate(scorer = case_when(points < 15 ~ 'low',
                           points < 25 ~ 'med',
                           points < 35 ~ 'high'))

# 2: Create New Variable Based on Several Existing Variables
#define new variable 'type' using mutate() and case_when()
df %>%
  mutate(type = case_when(player == 'a' | player == 'b' ~ 'starter',
                            player == 'c' | player == 'd' ~ 'backup',
                            position == 'G' ~ 'reserve'))

# create a new variable called ‘valueAdded’ based on the value in the points and rebounds columns
#define new variable 'valueAdded' using mutate() and case_when()
df %>%
  mutate(valueAdded = case_when(points <= 15 & rebounds <=5 ~ 2,
                                points <=15 & rebounds > 5 ~ 4,
                                points < 25 & rebounds < 8 ~ 6,
                                points < 25 & rebounds > 8 ~ 7,
                                points >=25 ~ 9))

```


# tidyr  
<br>A package designed for creating tidy data.

## Spread Function
<br>**spread(data, key value)**
<br>where,  *data:* Name of the data frame
<br<*key:* Column whose values will become variable names
<br>*value:* Column where values will fill under new variables created from key
<br>1. *Spread Values Across Two Columns*
<br>2. *Spread Values Across More Than Two Columns*

```{r spread1}
library(tidyr)
#create data frame
df <- data.frame(player=rep(c('A', 'B'), each=4),
                 year=rep(c(1, 1, 2, 2), times=2),
                 stat=rep(c('points', 'assists'), times=4),
                 amount=c(14, 6, 18, 7, 22, 9, 38, 4))

#view data frame
df

# 1: Spread Values Across Two Columns
# to turn the values in the stat column into their own columns
#spread stat column across multiple columns
spread(df, key=stat, value=amount)

# 2: Spread Values Across More Than Two Columns
#create data frame
df2 <- data.frame(player=rep(c('A'), times=8),
                 year=rep(c(1, 2), each=4),
                 stat=rep(c('points', 'assists', 'steals', 'blocks'), times=2),
                 amount=c(14, 6, 2, 1, 29, 9, 3, 4))

#view data frame
df2

# turn the four unique values in the stat column into four new columns:
#spread stat column across multiple columns
spread(df2, key=stat, value=amount)
```

## Gather Function
<br>The gather() function from the tidyr package can be used to “gather” a key-value pair across multiple columns.
<br>**gather(data, key value, …)**
<br>where, *data:* Name of the data frame
<br>*key:* Name of the key column to create
<br>*value:* Name of the value column to create
<br>*… :* Specify which columns to gather from
<br>1. *Gather Values From Two Columns*
<br>2. *Gather Values From More Than Two Columns*

```{r gather1}
# 1: Gather Values From Two Columns
#create data frame
df <- data.frame(player=c('A', 'B', 'C', 'D'),
                 year1=c(12, 15, 19, 19),
                 year2=c(22, 29, 18, 12))

#view data frame
df

# Create two new columns called “year” and “points” as follows:
#gather data from columns 2 and 3
gather(df, key="year", value="points", 2:3)

# 2: Gather Values From More Than Two Columns
#create data frame
df2 <- data.frame(player=c('A', 'B', 'C', 'D'),
                  year1=c(12, 15, 19, 19),
                  year2=c(22, 29, 18, 12),
                  year3=c(17, 17, 22, 25))

#view data frame
df2

# “gather” the values from columns 2, 3, and 4 into two new columns called “year” and “points” as follows
#gather data from columns 2, 3, and 4
gather(df2, key="year", value="points", 2:4)
```

### Use Separate Function in R 
<br>The separate() function from the tidyr package can be used to separate a data frame column into multiple columns.
<br>**separate(data, col, into, sep)**
<br>where, **data:** Name of the data frame
<br>**col:** Name of the column to separate
<br>**into:** Vector of names for the column to be separated into
<br>**sep:** The value to separate the column at
<br>1. *Separate Column into Two Columns*
<br>2. *Separate Column into More Than Two Columns*

```{r seperte1}
# 1: Separate Column into Two Columns
#create data frame
df <- data.frame(player=c('A', 'A', 'B', 'B', 'C', 'C'),
                 year=c(1, 2, 1, 2, 1, 2),
                 stats=c('22-2', '29-3', '18-6', '11-8', '12-5', '19-2'))

#view data frame
df

#separate stats column into points and assists columns
separate(df, col=stats, into=c('points', 'assists'), sep='-')

# 2: Separate Column into More Than Two Columns
#create data frame
df2 <- data.frame(player=c('A', 'A', 'B', 'B', 'C', 'C'),
                 year=c(1, 2, 1, 2, 1, 2),
                 stats=c('22/2/3', '29/3/4', '18/6/7', '11/1/2', '12/1/1', '19/2/4'))

#view data frame
df2
# separate stats column into three new columns
separate(df2, col=stats, into=c('points', 'assists', 'steals'), sep='/')
```


## Unite Function 
<br>The unite() function from the tidyr package can be used to unite multiple data frame columns into a single column.
<br>**unite(data, col, into, sep)**
<br>where, **data:** Name of the data frame
<br>**col:** Name of the new united column
<br>**… :** Vector of names for the columns to unite
<br>**sep:** How to join the data in the new united column
<br>1. *Unite Two Columns into One Column*
<br>2. *Unite More Than Two Columns*

```{r unit1}
#  1: Unite Two Columns into One Column
#create data frame
df <- data.frame(player=c('A', 'A', 'B', 'B', 'C', 'C'),
                 year=c(1, 2, 1, 2, 1, 2),
                 points=c(22, 29, 18, 11, 12, 19),
                 assists=c(2, 3, 6, 8, 5, 2))

#view data frame
df

#unite points and assists columns into single column
unite(df, col='points-assists', c('points', 'assists'), sep='-')

# 2: Unite More Than Two Columns
#create data frame
df2 <- data.frame(player=c('A', 'A', 'B', 'B', 'C', 'C'),
                  year=c(1, 2, 1, 2, 1, 2),
                  points=c(22, 29, 18, 11, 12, 19),
                  assists=c(2, 3, 6, 8, 5, 2),
                  blocks=c(2, 3, 3, 2, 1, 0))

#view data frame
df2

#unite points, assists, and blocks column into single column
unite(df2, col='stats', c('points', 'assists', 'blocks'), sep='/')
```

https://www.statology.org/in-operator-in-r/